The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-10-17 poster/main.tex
Added short slides

Training large language models requires optimizers more scalable and efficient than Adam/AdamW. The recent Muon~[2] optimizer shows great promise, sometimes doubling AdamW's efficiency, due to its unique design:
\begin{itemize}
  \item \textbf{\color{HazySummerEve}Matrix-Aware Design:} It is constructed specifically for weight matrices rather than generic vectors.
  \item \textbf{\color{HazySummerEve}Principled Update Rule:} Muon's update is not merely heuristic. It is derived as the solution to a linear minimization problem constrained by the operator (spectral) norm, giving it a solid mathematical foundation.
\end{itemize}

Muon's success has inspired variants like Dion~[1] algorithm and frameworks of Scion and Gluon~[4]. We build on this work by asking a fundamental question:
\begin{attention}
In deriving Muon's update step, why constrain by the spectral norm? How would alternative norms affect performance and computational cost?
\end{attention}
To investigate this, we first analyze the link between Muon-like algorithms and their defining norms. We then propose \textbf{\color{HazySummerEve}F-Fanions}, a novel family of optimizers derived from exotic, mixed norms. Finally, we present an empirical comparison between Muon and F-Fanions.

\vspace{0.5em}
\textbf{\Huge\color{Zen}How Norms Shape the Update Step}\\[0.3em]

@@

{
\centering
% Using 'tabular' now for better width control based on content
\begin{tabular}{clcc}
\toprule
\textbf{Case} & \textbf{Method} & \textbf{Norm} & \textbf{Update Formula} \\
\midrule
\multirow{3}{*}{\rotatebox[origin=c]{90}{\parbox{1.5cm}{\centering Vec.}}}
& Normalized SGD    & $\ell_2$          & $-\eta g^t / \norm{g^t}_2$ \\
& SignSGD           & $\ell_\infty$     & $-\eta\sign(g^t)$ \\
& Gauss-Southwell CD    & $\ell_1$      & $-\eta \sum_{i \in \argmax|g_i^t|} \sign(g_i^t)e_i$ \\
\midrule
\multirow{3}{*}{\rotatebox[origin=c]{90}{\parbox{1.5cm}{\centering Mat.}}}
& Normalized SGD    & $\normf{\cdot}$   & $-\eta\mG^t / \normf{\mG^t}$ \\
& Muon     & $\norms{\cdot}$   & $-\eta\mU \mV^\top$ \\
& Dion (without EF)   & $\norm{\cdot}_{KF-k}^\dag$ & $-\eta\mU_k \mV_k^\top$ \\
\bottomrule
\end{tabular}\eqspace
}
Here $\norm{\cdot}_{KF-k}$ denotes Ky Fan $k$ norm, $\norm{\cdot}^\dag$ is the dual norm.

@@

This framework recovers existing algorithms as special cases:
\begin{itemize}
    \item \textbf{\color{HazySummerEve}Normalized SGD:} $\alpha = 0$.
    \item \textbf{\color{HazySummerEve}Dion} (without error feedback): $\alpha = 1$, $k < \min\{m, n\}$.
    \item \textbf{\color{HazySummerEve}Muon:} $\alpha = 1$, $k = \min\{m, n\}$.
\end{itemize}
We also identify and analyze \textbf{\color{HazySummerEve}F-Neon}, a computationally efficient member of this family where $k=1$.

@@

We benchmarked our method against Muon on the \texttt{modded nanogpt 2024} speedrun, a pretraining of a real LLM. Below are the results after 1750 steps,

{
\centering
\begin{tabular}{lccc}
\toprule
\textbf{Method} & \textbf{LR} & \textbf{Momentum} & \textbf{Final Loss} \\
\midrule
Muon (Baseline) & 0.05 & 0.95 & 3.279 (Pass) \\
F-Muon ($\alpha=0.5$) & 0.07 & 0.95 & 3.281 (Near Miss) \\
\bottomrule
\end{tabular}
\par
}

@@

We evaluate F-Muon against Muon by training a Convolutional Neural Network (CNN) on the CIFAR-10 Airbench. All results are averaged over multiple runs after 8 epochs.

@@

\begin{columns}[T,totalwidth=\textwidth]
  \begin{column}{0.47\textwidth}
    \setlength{\parskip}{0.7em}
\setlength{\parindent}{0pt}
\Large
    We can see that Properly tuned F-Muon can perform on par with Muon, achieving the same 94\% accuracy.

    Another Interesting observation here is that tuned F-Muon has much larger trust region than Muon, and thus makes larger step sizes (figure to the right).
  \end{column}
  \begin{column}{0.49\textwidth}
    \includegraphics[width=\linewidth]{fstardual_cifar.pdf}
    \centering
  \end{column}
\end{columns}

@@

\begin{itemize}
    \item We introduced the \textbf{\color{HazySummerEve}F-Fanion framework}, unifying LMO-based optimizers like Muon and Normalized SGD via mixed norms.

    \item The choice of norm is a flexible design parameter, not a fixed rule. Our results show alternatives to the spectral norm can be equally effective.

    \item This work opens the core question of how to theoretically guide norm selection. Future work could also explore adapting the norm (via $\alpha$ and $k$) dynamically during training.
\end{itemize}

>>>>

LLM training requires more scalable and efficient optimizers. The recent Muon~[2] sometimes doubles AdamW's efficiency due to Muon's unique design :
\begin{itemize}
  \item \textbf{\color{HazySummerEve}Matrix-Aware Design:} Muon is constructed specifically for weight matrices rather than for generic vectors.
  \item \textbf{\color{HazySummerEve}Theory-Based Update Rule:} Muon's update is not a mere heuristic. It is derived as the solution to a linear minimization problem constrained by the operator (spectral) norm, giving it a solid mathematical foundation.
\end{itemize}

Muon's success has inspired variants like Dion~[1] algorithm and frameworks of Scion and Gluon~[4]. We build on this work by asking a fundamental question:
\begin{attention}
Why should the choice of a matrix norm in the Muon’s update derivation be restricted to the spectral norm? How might other matrix norms influence the algorithm’s performance and computational cost?
\end{attention}
To investigate this, we first analyze the link between Muon-like algorithms and their defining norms. We then propose \textbf{\color{HazySummerEve}F-Fanions}, a novel family of optimizers derived from exotic, mixed norms. Finally, we present an empirical comparison between Muon and F-Fanions.

\vspace{0.5em}
\textbf{\Huge\color{Zen}How Norms Shape the Update}\\[0.3em]

@@

{
\centering
% Using 'tabular' now for better width control based on content
\begin{tabular}{clcc}
\toprule
\textbf{Case} & \textbf{Method} & \textbf{Norm} & \textbf{Update Formula} \\
\midrule
\multirow{3}{*}{\rotatebox[origin=c]{90}{\parbox{1.5cm}{\centering Vec.}}}
& Normalized SGD    & $\ell_2$          & $-\eta g^t / \norm{g^t}_2$ \\
& SignSGD           & $\ell_\infty$     & $-\eta\sign(g^t)$ \\
& Gauss-Southwell CD    & $\ell_1$      & $-\eta \sum_{i \in \argmax|g_i^t|} \sign(g_i^t)e_i$ \\
\midrule
\multirow{4}{*}{\rotatebox[origin=c]{90}{\parbox{1.5cm}{\centering Mat.}}}
& Normalized SGD    & $\normf{\cdot}$   & $-\eta\mG^t / \normf{\mG^t}$ \\
& Muon     & $\norms{\cdot}$   & $-\eta\mU \mV^\top$ \\
& Neon (ours)   & $\norm{\cdot}_{nuc}$ $ & $-\eta u_1 v_1^\top$ \\
& Dion (without EF)   & $\norm{\cdot}_{KF-k}^\dag$ & $-\eta\mU_k \mV_k^\top$ \\

\bottomrule
\end{tabular}\eqspace
}
Here $\norm{\cdot}_{KF-k}$ denotes Ky Fan $k$ norm, $\norm{\cdot}^\dag$ is the dual norm.

@@

This framework recovers existing algorithms as special cases:
\begin{itemize}
    \item \textbf{\color{HazySummerEve}Normalized SGD:} $\alpha = 0$.
    \item \textbf{\color{HazySummerEve}Dion} (without error feedback): $\alpha = 1$, $k < \min\{m, n\}$.
    \item \textbf{\color{HazySummerEve}Muon:} $\alpha = 1$, $k = \min\{m, n\}$.
\end{itemize}
We also identify and analyze \textbf{\color{HazySummerEve}F-Neon}, a computationally efficient member of this family with $k=1$.

@@

We benchmarked our method against Muon and NSGD on the \texttt{modded nanogpt 2024} speedrun, which is pretraining of a real LLM. Below are the results after 1,750 steps. The objective was to achieve a loss lower than 3.28.

{
\centering
\begin{tabular}{lccc}
\toprule
\textbf{Method} & \textbf{LR} & \textbf{Momentum} & \textbf{Final Loss} \\
\midrule
Muon (Baseline)         & 0.05 & 0.95 & 3.279  \\
F-Muon ($\alpha=0.5$)   & 0.07 & 0.95 & 3.281 \\
NSGD                    & 0.07 & 0.96 & 3.4651 \\
F-Muon ($\alpha=0.5$)   & 0.07 & 0.96 & 3.2824 \\
\bottomrule
\end{tabular}
\par
}

@@

We evaluate F-Muon against Muon by training a Convolutional Neural Network (CNN) on the CIFAR-10 Airbench. All results are averaged over multiple runs and reported after 8 epochs with batch size = 2,000.

@@

\begin{columns}[T,totalwidth=\textwidth]
  \begin{column}{0.47\textwidth}
    \setlength{\parskip}{0.7em}
\setlength{\parindent}{0pt}
\Large
    We observe that a properly tuned F-Muon performs on par with Muon and achieves the same 94.01\% accuracy.

    Moreover, the tuned F-Muon has a significantly larger trust region than Muon, and the learning rate is so high that Muon yields lower test accuracy when trained with it (see the figure on the right).
  \end{column}
  \begin{column}{0.49\textwidth}
    \includegraphics[width=\linewidth]{fstardual_cifar.pdf}
    \centering
  \end{column}
\end{columns}

@@

\begin{itemize}
    \item We introduced the \textbf{\color{HazySummerEve}F-Fanions} that unify LMO-based optimizers like Muon and Normalized SGD via mixed norms.

    \item The choice of a norm is a flexible design parameter rather than a fixed rule. Our results show that other norms may be as efficient as the spectral norm.

    \item Our work poses a crucial question of how to theoretically guide norm selection. Future work could also explore adapting the norm (via $\alpha$ and $k$) dynamically during training.
\end{itemize}
